Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-09-04 Paper.qmd
Updated plots to better show the effect of predictors.
Added to limitations section and edited the introduction, condensing and removing sections.

---
title: "Child Signals of Need in Utila, Honduras"
execute: 
  echo: false
  warning: false
format: html
lightbox: true
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
---

@@

We argue that older children (i.e., those older than infants) have evolved strategies to tailor their signals to 1) their valuation of a given resource, 2) the extent of conflicts of interest with caretakers, and 3) the presence of existing information about a child's need or potential to benefit from a resource so that they can credibly convey they are worth additional investment. Such strategies can lead to some combination of verbal requests for more support, facial sadness, crying, withdrawal, temper tantrums and self-harm, all of which may be effective in certain circumstances and are sometimes successful in eliciting more support from caretakers, despite a child's low status and formidability.

@@

Throughout human evolutionary history, individuals often found themselves in situations in which they received fitness-relevant benefits from others that could not easily be replaced [@apicella2012; @apicella2019; @hill2011; @tooby1996]. In these instances, natural selection is expected to have favored those who provided support to others if 1) such help meant one could expect to receive fitness benefits from those helped in the future and 2) the costs of providing help were outweighed by the benefits to the helper on average[@andré2010; @hamilton1964; @tooby1996; @trivers1971]. At the very least, this would have taken the form of mothers providing investment to their children due to direct fitness benefits.

The fitness interdependence of offspring and at least one parent is seen in many animals [@clutton-brock1991; @webb1999]. However, humans are a uniquely interdependent species among the living primates [@tomasello2012], with one’s fitness prospects depending greatly on a network of kin and other social allies [@apicella2019]. Although also true with adults, this need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities[@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from parents[@kramer2018], with most human reproduction occurring in the context of biparental investment and long-term pair bonds [@apostolou2007; @schacht2019]. However, in many societies, other adults like grandparents, aunts, uncles, and non-relatives have also been found to provide significant investment in children [@kramer2018; @sear2008], something which was likely favored due to indirect fitness benefits, reciprocity, or reputation management [@gurven2000] + (Schniter Preprint (permission needed). Children also often serve as alloparents, helping to raise their younger siblings or provide for them through food preparation or collection [@hrdy2009; @meehan2009; @sear2008], despite still requiring substantial amounts of investment themselves [@kramer2018a].

@@

Conflicts of interest within families are largely structured by the likelihood an individual shares their genes with another. As a sexually reproducing species, humans have a 50% chance of passing down a given gene to each child. Each child, however, shares more genes with themselves (100%) than with their full siblings (50% on average), half-siblings (25% on average), and stepsiblings (0%). Due to this asymmetry, children are often expected to attempt to elicit more investment from parents than is optimal from the perspective of the parent, while parents are expected to attempt to invest less than is optimal for each child in order to invest in their other children, whom parents are expected to value identically, all else being equal [@trivers1974].

For the same reason, siblings also often find themselves in conflicts of interest with one another, with each being expected to prefer more resources for themself than others. Sibling competition is evident across taxa and can range from relatively benign forms like strong begging competition in chicks [@caro2016] to siblicide in birds, mammals, and sharks [@salmon2014]. In humans, sibling competition is largely estimated by counts of siblings, with large numbers of siblings having been associated with decreases in growth [@hadley2011; @hagen2006; @salmon2014], survival [@lawson2012; @meij2009; @strassmann2011], and reproductive success [@gibson2011; @mace1996] in multiple populations, highlighting the likely significance of sibling competition for resources, even if this does not always manifest with behavioral conflict [@bateson1994].

While parent-offspring conflict and sibling competition are likely ubiquitous, the resources being contested will vary across human groups. In relatively egalitarian and food constrained populations, conflict over resources is likely to be centered on food, which probably explains the negative impacts siblings have on child age-specific heights and weights among the Shuar of Ecuador [@hagen2009], although similar effects have also been found in industrialized populations [@lawson2008]. In societies with greater inequality, competition is more likely to be over wealth, as is the case with older brothers in pastoralist communities like the Gabbra and Arsi Oromo limiting the marriage prospects of laterborn sons [@gibson2011; @mace1996]. Compared to large-scale investments like bride price or inheritance, little is understood about smaller transfers of wealth to younger children despite the potential importance of such transfers to 1) facilitate the development of embodied capital outside of growth and fat reserves or 2) help children succeed in their peer groups.

@@

Infant cries have been found to result in increased amounts of care, resources, and attention they receive and increased proximity to parents [@lin2012]. For example, Karo Batak infants who cried more were breastfed more, and the length of crying bouts predicted the time mothers breastfed [@kushnick2006]. Similarly, !Kung San mothers frequently respond to crying infants by providing comfort and breastfeeding access, with comforting behaviors, in general, being far more common than one’s meant to restrict infant behavior [@kruger2010]. Although this study found mothers were the most common responders, siblings, fathers, and other adults frequently stepped in so that 88% of crying bouts were responded to, with most of the cries not responded to lasting under ten seconds [@kruger2010]. Similar responsiveness has also been reported among the Aka [@hewlett1998], Mbendjele hunter-gatherers [@chaudhary2024], the Efe [@tronick1987], and the Ache, among whom young children have been described as “win\[ning\] most conflicts with parents simply by crying and whining” [@kaplan1987]. Studies examining the potential for long-term benefits are rare and difficult to achieve; however, during a time of high infant mortality in a sample of 13 Massai infants, those with easy temperaments were more likely to die (5/7) than those with difficult temperaments (1/6), something which is consistent with signals resulting in beneficial treatment [@devries1987].

Crying does not always succeed in eliciting benefits, however. Compared to the previously mentioned groups, the Ngandu [@hewlett1998] and the Hadza [@blurtonjones1993] have been found to be more likely to ignore the cries of infants and young children, and concerns over the risk of spoiling children by being too responsive seem to be reasonably common in some Western populations [@barnett2010; @burchinal2010; @chaudhary2024; @frankenhuis2022; @maute2018]. Crying can also lead to negative responses beyond being ignored [@barr1990; @soltis2004; @zeifman2017]. It has been found to increase parental distress, the likelihood of maternal depression, and the child’s risk of maltreatment ~~especially if the child is in poor condition~~ [@frodi1985; @fujiwara2011; @howard2006; @lin2012; @soltis2004]. For example, in a sample of 3,259 parents of Dutch infants under 6 months old, 5.6% self-reported smothering, slapping, or shaking their babies in response to their crying, outcomes which were more likely among parents who were single, unemployed, or deemed their infant’s crying to be excessive [@reijneveld2004]. Excessive infant crying, i.e., colic, has also been found to be a risk factor for multiple negative outcomes [@zeevenhooven2018], including abuse, shaken baby syndrome [@kurth2011], and earlier cessation of breastfeeding [@howard2006].

Although widely accepted as a signal, there is no consensus on why emotional crying may work to elicit positive responses. One possibility is that crying is physiologically constrained in ways that prevent or limit faking. Evidence consistent with this comes from findings that crying is hard to do on command and findings individuals often cry while alone, something which is not predicted by signaling theory [@vingerhoets2001]. However, the likelihood of crying is often clearly related to presence of supporters, something which is seen in findings that lonely individuals cry less than those with more support options despite reporting lower well-being [@vingerhoets2013] and the existence of delayed crying—times when crying is inhibited during the cause of negative affect only for it to be expressed when one is near potential supporters [@gracanin2018]. A more likely possibility is that certain acoustic elements of infant cries are tied to their physiological state in ways that make faking certain states impossible, with crying serving as a graded signal in most other instances [@soltis2004; @zeifman2017].

@@

Withdrawal has been suggested to be an underappreciated way those with low social power may credibly signal need [@syme2023a]. Due to their lower status, children have limited ways to change how others treat them. However, as children 1) are their parents’ path to direct fitness and 2) can help family members through allocare and other forms of labor, they provide benefits that are valuable and hard to replace. By withholding these benefits or putting the future of these benefits into question, a child simultaneously delays the benefits they get from this cooperation while imposing costs on those with a fitness interest in the child. As these costs are likely lower for those truly in need than those who are not, withdrawal may be a credible way to signal need if the costs to faking the signal make it unprofitable. In these cases, the end result is analogous to a labor strike, in which the signaler conveys to others their situation demands more support, and if they want to continue receiving benefits from the child, they need to upregulate their support of the child.

@@

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to indicate need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010]. However, facial expressions, along with tears, may not be enough to credibly signal need in times of significant conflicts of interest. In these instances, depression and suicidal behavior have been suggested to function as ways to credibly signal need due to the expectation that such states are costlier for those facing adversity than those who are not (for more detail on these approaches see: [@hagen2003; @rosenthal1993; @syme2016; @watson2002].

@@

Another possibility is that temper tantrums carry the risk or implication of offspring self-harm. For example, Utilian mothers reported in focus groups that some temper tantrums involved children banging their heads on things when they do not get what they want, something that has been reported in other populations [@belden2008; @burrows1953; @daniels2012; @dennis1940; @vandenakker2022]. Similarly, young chimpanzees (*Pan troglodytes*) have been observed to beat their heads on the ground when refused access to breastmilk [@goodall1986], and American white pelican chicks (*Pelecanus erythrorhynchos*) have been found to throw themselves on the ground, flail violently, and bite at their own wings when begging for food [@schaller1964]. Such behaviors have been suggested to follow similar logic to depression and suicide attempts as they are likely relatively less costly to those who are truly in dire need compared to those who are not to the point faking is not worth the costs [@hagen2008] (Hagen et al. 2008). In other words, the risk of injury from violent temper tantrums carries greater fitness costs to a child who is receiving enough support to develop high levels of embodied capital than it does to a child who is neglected to the point where their survival prospects are low. In this scenario, the risk of harm is likely not worth the potential benefits of increased support to the child receiving substantial investment, but it may be for the neglected child who may have few other options to compel others to invest in them.

@@

This study was designed to help establish the natural history of child signaling behavior in Utila, Honduras, by quantifying the frequencies of common child signals of need and various types of parental responses. From this, we sought to examine 1) the causes of child signals of need, 2) how and why signal costs might vary, 3) how parents respond to indications of child need, 4) how signaling behavior changes with age and sex, 5) and if there were differences in child signaling and parental responses among two sub-populations.

A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. This hypothisized causal, bi-directional relationship between investment and signaling makes it difficult to isolate the effects of signaling behavior on investment in children when using standard measures of child welfare (i.e., BMI, health metrics, and height for age). When experiments are not appropriate, longitudinal studies with carefully selected variables can help address this (cite). However, our small sample size with signaling data and anthropometric data at two time periods prevents this for this study. For this reason, all analyses for this study were exploratory and meant to inform future studies of child signaling.

@@

Triceps and subscapular skinfold thicknesses were measured using skinfold calipers. Measurements were taken directly on the skin. The triceps measurement occurred at the midpoint of the upper arm. The subscapular measurement occured immediately below the inferior angle of the scapula, being picked up diagonally at the natural fold line of the skin [@antóna].

@@

@@

@@

Alluvial plot showing the combinations of signaling frequencies. Alluvium width equals the number of individiuals with the same frequencies for the three signals/states. Stratum height equals the count of individuals at a given frequency. Frequency key: Low includes "never" and "once a month or less," mediums includes "more than once a month but less than once a week" and "more than once a week but not daily," high includes "daily," and "multiple times per day."
:::

@@

Stacked bar plot showing the frequency of conflict between child and primary caregiver as reported by primary caretaker. Color corresponds to sex of the child. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day."
:::

@@

Stacked bar plot showing the frequency of conflict a child's alloparenting effort as reported by primary caretaker. Color corresponds to sex of the child. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day."
:::

@@

The effect of age on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-main_models for the corresponding regression table.
:::

@@

The effect of the frequency of conflict with one's primary caretaker on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-main_models for the corresponding regression table.
:::

@@

::: {#fig-effect_edu}
```{r}
signaling_data_plot_edu
```

@@

The effect of A) child age and B) the number of adults in the household that contribute to childrcare on the frequency of conflict between caretaker and child. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-conflict_dep_var for the corresponding regression table.
:::

@@

#FIX coords do not work for this

@@

The effect of A) child age and B) the number of adults that contribute to childrecare within the household on the frequency of conflict between child and caretaker. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of child alloparenting frequency with a small amount of jitter to reveal overlapped values. See @tbl-conflict_dep_var for the corresponding regression table.
:::

@@

::: {#fig-effect_alloparenting_dep_var_plot}
```{r}
alloparenting_dep_var_plot
```

@@

# FIX: want y axes to extend from 0-1.00

@@

# FIX: want y axes to extend from 0-1.00

@@

Monthly income was also not found to be a significant predictor of signaling behavior. Many models designed to explain signals of need (e.g., the begging of young birds or withdrawal in humans) predict that offspring who are worse off or are receiving less investment have more to gain from signaling than those with more (cite). However, it is possible that this can be true and household level resources have little influence on child signaling frequencies between families. As signaling behavior is costly to child, caretakers, and siblings, children may gain direct and indirect fitness benefits from limiting their signaling behavior in times of resource stress (i.e., when the chances of gaining investment or the size of investment are small). Similarly, children in high resource families may also be motivated to signal at higher than expected frequencies due to the potential for larger pieces of investment. We know of no empirical explorations of this in humans (check more and confirm). However, multiple game theoretical models of sibling competition and have found that parental resources have a small impact on offspring signaling, especially with high rates of monogamy [Bossan + Mock and Parker + Check for newer sources]. Although counterintuitive, such an outcome fits the fact that fitness is relative and that natural selection should be expected to have favored those who maximized their own inclusive fitness, whether one found themselves in a great situation or a dire one.

@@

A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. In other words, higher levels of signaling may stem from greater levels of need while simultaneously eliciting more investment from others. This hypothisized causal, bi-directional relationship makes it difficult to isolate the effects of signaling behavior on investment in children when using standard measures of child welfare (i.e., BMI, health metrics, and height for age). Longitudinal studies can help address this. However, our small sample size of participants with signaling data and anthropometric data at two time periods prevents this for this study.

Due to the fact the data used here were a part of survey which covered many topics and often multiple children, we sometimes faced a tradeoff between survey time and question quality. Asking adult participants about their relative investment in children in their household was likely not as effective as more in depth, questions focused on specific questions. Framing the question this way may have also increased the likelihood of social desirability bias [@grimm2010], which could lead adult participants to be be more likely to report equal investment across children. Our questions about the last time the child signaled may have also been more effective if they focused on specific signals due to our methods combining crying, which is unambiguous as a signal, with sadness, which has both signaling and non-signaling functions.

>>>>

---
title: "Child Signals of Need in Utila, Honduras: An Exploratory Study"
execute: 
  echo: false
  warning: false
format: html
lightbox: true
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
---

@@

In this paper, we use exploratory analysis on child signaling data from Utila Honduras to examine the hypothesis that children have evolved strategies to tailor their signals of need to 1) their valuation of a given resource, 2) the extent of conflicts of interest with caretakers, and 3) the presence of existing information about a child's need or potential to benefit from a resource so that they can credibly convey they are worth additional investment. Such strategies can lead to some combination of verbal requests for more support, facial sadness, crying, withdrawal, temper tantrums and self-harm, all of which may be effective in certain circumstances and are sometimes successful in eliciting more support from caretakers, despite a child's low status and formidability.

@@

Throughout human evolutionary history, individuals often found themselves in situations in which they received fitness-relevant benefits from others that could not easily be replaced [@apicella2012; @apicella2019; @hill2011; @tooby1996]. In these instances, natural selection is expected to have favored those who provided support to others if 1) such help meant one could expect to receive fitness benefits from those helped in the future and 2) the costs of providing help were outweighed by the benefits to the helper on average [@andré2010; @hamilton1964; @tooby1996; @trivers1971]. At its simplest, this would have taken the form of mothers providing investment to their children.

The fitness interdependence of offspring and at least one parent is seen in many animals [@clutton-brock1991; @webb1999]. However, humans are a uniquely interdependent species among the living primates [@tomasello2012] in that one’s fitness prospects depend greatly on a network of kin and other social allies [@apicella2019]. Although also true with adults, this need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities [@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from parents [@kramer2018], with most human reproduction occurring in the context of biparental investment and long-term pair bonds [@apostolou2007; @schacht2019]. However, in many societies, other adults like grandparents, aunts, uncles, and non-relatives have also been found to provide significant investment in children [@kramer2018; @sear2008], something which was likely favored due to indirect fitness benefits, reciprocity, or reputation management [@gurven2000] + (Schniter Preprint (permission needed). Children also often serve as alloparents, helping to raise their younger siblings or provide for them through food preparation or collection [@hrdy2009; @meehan2009; @sear2008], despite still requiring substantial amounts of investment themselves [@kramer2018a].

@@

Conflicts of interest within families are largely structured by the likelihood an individual shares their genes with another. As a sexually reproducing species, humans have a 50% chance of passing down a given gene to each child. Each child, however, shares more genes through descent with themselves (100%) than with their full siblings (50% on average), half-siblings (25% on average), and stepsiblings (0%). Due to this asymmetry, children are often expected to attempt to elicit more investment from parents than is optimal from the perspective of the parent, while parents are expected to attempt to invest less than is optimal for each child in order to invest in their other children, whom parents are expected to value identically, all else being equal [@trivers1974].

For the same reason, siblings also often find themselves in conflicts of interest with one another, with each being expected to prefer more resources for themself than others. Sibling competition is evident across taxa and can range from relatively benign forms like strong begging competition in chicks [@caro2016] to siblicide in birds, mammals, and sharks [@salmon2014]. In humans, sibling competition is largely estimated by counts of siblings, and large numbers of siblings have been associated with decreased growth [@hadley2011; @hagen2006; @salmon2014], survival [@lawson2012; @meij2009; @strassmann2011], and reproductive success [@gibson2011; @mace1996] in multiple populations, highlighting the likely significance of sibling competition for resources, even if this does not always manifest with behavioral conflict [@bateson1994].

Potentially cut this paragraph: While parent-offspring conflict and sibling competition are likely ubiquitous, the resources being contested will vary across human groups. In relatively egalitarian and food constrained populations, conflict over resources is likely to be centered on food, which probably explains the negative impacts siblings have on child age-specific heights and weights among the Shuar of Ecuador [@hagen2009], although similar effects have also been found in industrialized populations [@lawson2008]. In societies with greater inequality, competition is more likely to be over wealth, as is the case with older brothers in pastoralist communities like the Gabbra and Arsi Oromo limiting the marriage prospects of laterborn sons [@gibson2011; @mace1996]. Compared to large-scale investments like bride price or inheritance, little is understood about smaller transfers of wealth to younger children despite the potential importance of such transfers to 1) facilitate the development of embodied capital outside of growth and fat reserves or 2) help children succeed in their peer groups.

Conflicts of interest between children and parents may also involve how much children invest in the household. As previously mentioned, each individual child is expected to prefer greater investment in oneself than in one’s siblings, while parents, who are equally related to all their children, should often be expected to favor a more even distribution [@trivers1974]. This can lead to conflict with older children who might prefer to spend their time in ways which are likely to result in future direct fitness benefits (e.g., finding romantic partners or forming strong bonds with friends) rather than taking care of their siblings or acquiring or preparing food [@lowe2003; @syme2023a]. Conflicts of interest over investment in the household also appear to extend to when a child starts having their own children [@kushnick2023; @emlen1995], with parents potentially benefiting from delaying the reproductive careers of certain offspring in order to gain more care for their other children [@flinn1989; @emlen1992] or speeding up a child’s departure from the household if the child’s presence limits how much caretakers can invest in other children they deem more worthy of investment [@kushnick2023].

@@

Infant cries have been found to result in increased amounts of care, resources, and attention they receive and increased proximity to parents [@lin2012]. For example, Karo Batak infants who cried more were breastfed more, and the length of crying bouts predicted the time mothers breastfed [@kushnick2006]. Similarly, !Kung San mothers frequently respond to crying infants by providing comfort and breastfeeding access, with comforting behaviors, in general, being far more common than one’s meant to restrict infant behavior [@kruger2010]. Although this study found mothers were the most common responders, siblings, fathers, and other adults frequently stepped in so that 88% of crying bouts were responded to, with most of the cries not responded to lasting under ten seconds. Similar responsiveness has also been reported among the Aka [@hewlett1998], Mbendjele hunter-gatherers [@chaudhary2024], the Efe [@tronick1987], and the Ache, among whom young children have been described as “win\[ning\] most conflicts with parents simply by crying and whining” [@kaplan1987]. Studies examining the potential for long-term benefits are rare and difficult to achieve; however, during a time of high infant mortality in a sample of 13 Massai infants, those with easy temperaments were more likely to die (5/7) than those with difficult temperaments (1/6), something which is consistent with signals resulting in beneficial treatment [@devries1987].

Crying does not always succeed in eliciting benefits, however. Compared to the previously mentioned groups, the Ngandu [@hewlett1998] and the Hadza [@blurtonjones1993] have been found to be more likely to ignore the cries of infants and young children, and concerns over the risk of spoiling children by being too responsive seem to be reasonably common in some Western populations [@barnett2010; @burchinal2010; @chaudhary2024; @frankenhuis2022; @maute2018]. Crying can also lead to negative responses beyond being ignored [@barr1990; @soltis2004; @zeifman2017]. It has been found to increase parental distress, the likelihood of maternal depression, and the child’s risk of maltreatment [@frodi1985; @fujiwara2011; @howard2006; @lin2012; @soltis2004]. For example, in a sample of 3,259 parents of Dutch infants under 6 months old, 5.6% self-reported smothering, slapping, or shaking their babies in response to their crying, outcomes which were more likely among parents who were single, unemployed, or deemed their infant’s crying to be excessive [@reijneveld2004]. Excessive infant crying, i.e., colic, has also been found to be a risk factor for multiple negative outcomes [@zeevenhooven2018], including abuse, shaken baby syndrome [@kurth2011], and earlier cessation of breastfeeding [@howard2006].

Although widely accepted as a signal, there is no consensus on why emotional crying may work to elicit positive responses. One possibility is that the physiological costs of crying for newborns and infants [@rao1997; @thureen1998] may be prohibitively costly for faking to pay off. When this is the case, crying may serve as a credible of vigor [@furlow1997; @lummaa1998; @soltis2004; @wells2003], something which may be particularly important early in life due to increased risk of infanticide for children who have has poor survival prospects, due to either health conditions or resource constraints [@daly1984; @soltis2004].

Potentially combine and condense with paragraph below: As children age, their reproductive value increases due to reduced odds of mortality, leading to a decreased risk of infanticide (Hagen 2004). For this reason, there is likely a transition period in which crying is primarily meant to provide information about how much a child values further investment. Many models have been put forward which attempt to explain human crying and begging in other species as credible signals of need due to the differential costs and benefits of crying and analogous vocalizations in other species as deterring the exaggeration of need [@godfray1991; @kilner1997; @parker2002]. Although differing in their details, a common theme is that the indirect fitness costs signalers incur from imposing costs on that parents and siblings through begging (be them energetic costs, survival costs, or reduced likelihood of future siblings) creates a floor to possible deception but falls short of preventing it [@daly1990; @wells2003].

For older children, the relatively insignificant energetic costs of crying may not be enough to result in credible signals in times of substantial conflicts of interest. For this reason, crying among older children may be more similar to adult crying, something which has been found to result in increased support and empathy but sometimes come at the cost of reputation and the potential for stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs may be a means by which crying can still be a credible signal [@sznycer] with moderate conflicts of interest if they dissuade potential fakers from making the signal.

@@

Withdrawal has been suggested to be an underappreciated way those with low social power may credibly signal need [@syme2023a]. Due to their lower status, children have limited ways to change how others treat them. However, as children 1) are a major component of parents' inclusive fitness and 2) can help family members through allocare and other forms of labor, they provide benefits that are valuable and hard to replace. By withholding these benefits or putting the future of these benefits into question, a child simultaneously delays the benefits they get from this cooperation while imposing costs on those with a fitness interest in the child. As these costs are likely lower for those truly in need than those who are not, withdrawal may be a credible way to signal need if the costs to faking the signal make it unprofitable. In these cases, the end result is analogous to a labor strike, in which the signaler conveys to others their situation demands more support, and if they want to continue receiving benefits from the child, they need to upregulate their support of the child.

@@

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to indicate need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010]. However, facial expressions, along with tears, may not be enough to credibly signal need in times of significant conflicts of interest. In these instances, depression and suicidal behavior have been suggested to function as ways to credibly signal need due to the expectation that such states are costlier for those facing adversity than those who are not (for more detail on these approaches see: [@hagen2003; @rosenthal1993; @syme2016; @watson2002; @gaffney2022].

@@

Another possibility is that temper tantrums carry the risk or implication of offspring self-harm. For example, Utilian mothers reported in focus groups that some temper tantrums involved children banging their heads on things when they do not get what they want, something that has been reported in other populations [@belden2008; @burrows1953; @daniels2012; @dennis1940; @vandenakker2022]. Similarly, young chimpanzees (*Pan troglodytes*) have been observed to beat their heads on the ground when refused access to breast milk [@goodall1986], and American white pelican chicks (*Pelecanus erythrorhynchos*) have been found to throw themselves on the ground, flail violently, and bite at their own wings when begging for food [@schaller1964]. Such behaviors have been suggested to follow similar logic to depression and suicide attempts as they are likely relatively less costly to those who are truly in dire need compared to those who are not to the point faking is not worth the costs [@hagen2008]. In other words, the risk of injury from violent temper tantrums carries greater fitness costs to a child who is receiving enough support to develop high levels of embodied capital than it does to a child who is neglected to the point where their survival prospects are low. In this scenario, the risk of harm is likely not worth the potential benefits of increased support to the child receiving substantial investment, but it may be for the neglected child who may have few other options to compel others to invest in them.

@@

This study was designed to help establish the natural history of child signaling behavior in Utila, Honduras, by quantifying the frequencies of common child signals of need and various types of parental responses. From this, we sought to use signaling theory to examine 1) the causes of child signals of need, 2) how and why signal costs might vary, 3) how parents respond to indications of child need, 4) how signaling behavior changes with age and sex, 5) and if there were differences in child signaling and parental responses two groups who face different levels of social and economic adversity.

A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. This hypothisized causal, bi-directional relationship between investment and signaling makes it difficult to isolate the effects of signaling behavior on investment in children when using standard measures of child welfare (i.e., BMI, health metrics, and height for age). When experiments are not appropriate, longitudinal studies with carefully selected variables can help address this (cite). However, our small sample size with signaling data and anthropometric data at two time periods limits our ability to do this in this study. For this reason, all analyses for this study were exploratory and meant to inform future studies of child signaling.

@@

Triceps and subscapular skinfold thicknesses were measured using skinfold calipers. Measurements were taken directly on the skin. The triceps measurement occurred at the midpoint of the upper arm. The subscapular measurement occured immediately below the inferior angle of the scapula, being picked up diagonally at the natural fold line of the skin [@antóna]. Add flexed bicep measurementxxxxxx.

@@

To model the effects of skinfold thickness, flexed bicep circumference, and grip strength on signaling behavior, we used the residuals from generalized additive models which modeled the effect of sex and age on each measurement using the mgcv package [@wood2011]. This resulted in variables which represented how each child differed from the average child for their given sex and age.

@@

All statistical analyses were exploratory and involved fitting multiple models while being cognizant of the risk of overfitting the data, but with the payoff of idenftifying theoretically relavent effecrs that should be tested in the future.

@@

Alluvial plot showing the combinations of signaling frequencies per month. Alluvium width equals the number of individiuals with the same frequencies for the three signals/states. Stratum height equals the count of individuals at a given frequency. Frequency key: Low includes "never" and "once a month or less," mediums includes "more than once a month but less than once a week" and "more than once a week but not daily," high includes "daily," and "multiple times per day."
:::

@@

Stacked bar plot showing the frequency of conflict (per month) between child and primary caregiver as reported by primary caretaker. Color corresponds to sex of the child. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day."
:::

@@

Stacked bar plot showing the frequency of a child's alloparenting effort (per month) as reported by primary caretaker. Color corresponds to sex of the child. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day."
:::

@@

The effect of age on specific signal frequencies (per month), summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-main_models for the corresponding regression table.
:::

@@

The effect of the frequency of conflict with one's primary caretaker (per month) on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-main_models for the corresponding regression table.
:::

@@

::: {#fig-effect_edu}
```{r}
signaling_data_plot_education
```

@@

The effect of A) child age and B) the number of adults in the household that contribute to childcare on the frequency of conflict between caretaker and child. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-conflict_dep_var for the corresponding regression table.
:::

@@

@@

The effect of an alloparenting effort by sex interaction on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of child alloparenting frequency with a small amount of jitter to reveal overlapped values. See @tbl-conflict_dep_var for the corresponding regression table.
:::

@@

::: {#fig-effect_alloparenting_dep_var_plot}
```{r}
# alloparenting_dep_var_plot
```

@@

# FIX: error ribbons cross 1 and 0

@@

@@

Monthly income was also not found to be a significant predictor of signaling behavior. Many models designed to explain signals of need (e.g., the begging of young birds or withdrawal in humans) predict that offspring who are worse off or are receiving less investment have more to gain from signaling than those with more (cite). However, it is possible that this can be true and household level resources have little influence on child signaling frequencies between families. As signaling behavior is costly to child, caretakers, and siblings, children may gain direct and indirect fitness benefits from limiting their signaling behavior in times of resource stress (i.e., when the chances of gaining investment or the size of investment are small). Similarly, children in high resource families may also be motivated to signal at higher than expected frequencies due to the potential for larger pieces of investment. We know of no empirical explorations of this in humans (check more and confirm). However, multiple game theoretical models of sibling competition and have found that parental resources have a small impact on offspring signaling, especially with high rates of monogamy \[Bossan + Mock and Parker + Check for newer sources\]. Although counterintuitive, such an outcome fits the fact that fitness is relative and that natural selection should be expected to have favored those who maximized their own inclusive fitness, whether one found themselves in a great situation or a dire one.

@@

The main limitation of this study is that it is completely exploratory, something which results in a high risk of our models overfitting the data and indicating associations which might not exist in the real world. Our use of convenience and snowball sampling is also a limitation as it resulted in a non-probability sample which may not be representative of the broader population. This could happen if there was selection bias so that those who participated differed from the rest of the population in how they interacted with their children or how they assessed, remembered, and reported child behavior.

Outside of child anthropometric measurements, all data come from caregiver reports, and therefore, reflect the biases of caretakers, whatever they may be. Due to the fact the data used here were a part of survey which covered many topics and often multiple children, we also faced a tradeoff between survey time and question depth. Asking adult participants about their relative investment in children in their household was likely not as effective as more in depth, questions focused on specific questions. Framing the question this way may have also increased the likelihood of social desirability bias [@grimm2010], which could lead adult participants to be be more likely to report equal investment across children. Our questions about the last time the child signaled may have also been more effective if they focused on specific signals due to our methods combining crying, which is unambiguous as a signal, with sadness, which has both signaling and non-signaling functions.
